Predicting the 2024 US Presidential Election with a Model-Based Forecast

### 2024-11-04 paper/paper.qmd
update readme, finalize paper, fix reference

---
title: "Predicting the 2024 US Presidential Election with a Model-Based Forecast"
subtitle: "Using Generalized Linear Models to Predict Election Outcomes"
author: 
  - Yuanyi (Leo) Liu
  - Dezhen Chen
  - Ziyuan Shen
thanks: "Code and data are available at: [Forecasting the 2024 US Presidential Election](https://github.com/leoyliu/Forecasting-the-2024-US-Presidential-Election)."
date: today
date-format: long
abstract: "In this paper, we develop a linear model to predict the outcome of the 2024 U.S. presidential election using a \"poll-of-polls\" approach that aggregates data from multiple well-known pollsters. Our analysis projects that Donald Trump's probability of winning ranges from 30% to 52%, which depends on regional polling trends. However, Kamala Harris is projected to secure key states and obtain the majority of electoral votes, with a predicted count of 341 to Trump's 197 based on the testing dataset. These predictions are based on the testing dataset, which lacks data for some states, limiting our ability to produce results for all states. This paper underscores the effectiveness of aggregated polls in producing reliable election forecasts."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

The structure of the paper is organized as follows: following @sec-intro, @sec-data presents the data collection and cleaning process, along with an overview of the variables used in the analysis. @sec-model introduces the forecasting models, explaining why the selected models are suitable for predicting election outcomes based on aggregated polling data. Then @sec-result presents the main findings, including detailed crime trends for each neighborhood and year. Finally, @sec-discussion provides the results, highlighting key trends and predictions. Eventually, @sec-discussion concludes with a discussion of the findings, evaluating the reliability of the forecasts and identifying potential limitations of the models.

@@

ggplot(cleaned_data, aes(x = end_date, y = pct)) +
  geom_point() +  # Add points for individual poll data
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +  # Show month and year, with one label per month
  labs(
    x = "Date of Poll",
    y = "Polling Percentage (%)",
  ) +
  theme_minimal() +  # Use a clean minimal theme
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),  # Center and adjust title size
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for readability
  ) +
  geom_smooth()
```

@fig-trump-pct shows Donald Trump's polling percentages over time from July to October 2024. Each black dot represents an individual poll conducted on a specific date, indicating the percentage of voters who supported Trump in that poll. The blue line represents a trendline, smoothing out the individual poll results to show the general trajectory of Trump's support over time.

@@

Placeholder for a table showing the distribution of polling methodologies across different states.

@@

The model assumes that the distribution of support percentage, given these polling characteristics, follows a normal distribution. This Gaussian assumption facilitates parameter estimation, which is a standard approach in linear regression. To prevent overfitting and ensure interpretability, we assume moderate priors, maintaining balanced uncertainty across our predictors. This approach enables us to assess the impact of polling characteristics on candidate support while ensuring stability in our findings. Background details and diagnostics are included in the appendix.

@@

The linear regression model built by our training dataset, using 242 data points, estimated the factors influencing Trump's support levels. For brevity, @tbl-summary-model only shows the first ten rows of the model’s coefficients, with the full model summary available in the appendix. The intercept is estimated at 44.224, representing the baseline support level when all predictors are at their reference levels. The model achieved an $R^2$ value of 0.803, indicating that 80.3% of the variance in polling outcomes is explained by the predictors included in the model. The adjusted $R^2$ value of 0.753 confirms that the model captures the relationships between variables without significant overfitting.

@@

Other variables, like `transparency_score` (-0.220) and `duration` (0.028), have smaller effects. The negative coefficient for `transparency_score` suggests a slight decrease in predicted support as transparency increases, potentially reflecting a sensitivity among the public to the transparency level in polling methods. Meanwhile, `sample_size` (0.002) shows a small positive effect, indicating that larger sample sizes modestly increase predicted support. A complete overview of all predictors is provided in the appendix.

@@

# Create and display the results table
kable(missing_states, 
      col.names = c("State", "Electoral Votes", "Winner"), 
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  scroll_box(width = "100%", height = "400px")
```

Now, we combine the results from our predicted data and the historically inferred estimates for missing states. This merged dataset provides a view of the expected electoral outcomes across all states. @tbl-final-prediction presents the final predictions, showing the estimated electoral votes and projected winner for each state based on both model predictions and historical trends.

```{r}
#| label: tbl-final-prediction
#| tbl-cap: Final Predicted Electoral Outcomes by State, Combining Model Estimates and Historical Trends for Missing Data
#| echo: false
#| warning: false
#| message: false

# Remove the predicted_percent_trump column from state_predictions to match missing_states
state_predictions <- state_predictions %>%
  select(state, electoral_votes, winner)

# Merge the two datasets
final_predictions <- full_join(state_predictions, missing_states, by = c("state", "electoral_votes", "winner"))

# Arrange the rows by state name
final_predictions <- final_predictions %>%
  arrange(state)

# Display the final table
kable(final_predictions, 
      col.names = c("State", "Electoral Votes", "Winner"), 
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  scroll_box(width = "100%", height = "400px")
```

@@

@@

# Display the final table with added styling
kable(final_outcome, col.names = c("Candidate", "Final Votes"), digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  row_spec(1, background = "#FFCCCC") %>%    # Add light orange background for Trump row
  row_spec(2, background = "#C1D7FF") %>%    # Add light blue background for Harris row
  column_spec(2, bold = TRUE)
```

@@

# Plot the map using usmap and ggplot2
plot_usmap(data = final_predictions, values = "winner_color", regions = "states") +
  scale_fill_identity(
    name = "Predicted Winner",
    labels = c("Harris", "Trump"),
    guide = "legend"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 18),
    plot.background = element_rect(fill = "white", color = NA)
  )
```

@fig-statemap displays the predicted outcome for each state in the 2024 U.S. Presidential Election, with red representing states where Trump is expected to win and blue for those favoring Harris.

@@

# Model details {sec-model-details}
```{r}
#| label: fig-qq-check
#| fig-cap: Model Diagnosis
#| echo: false
#| warning: false
#| message: false

```

>>>>

---
title: "Predicting the 2024 US Presidential Election with a Model-Based Forecast"
subtitle: "Kamala Harris Predicted to Lead with 341 Electoral Votes Based on a Poll-Based Prediction"
author: 
  - Yuanyi (Leo) Liu
  - Dezhen Chen
  - Ziyuan Shen
thanks: "Code and data are available at: [Forecasting the 2024 US Presidential Election](https://github.com/leoyliu/Forecasting-the-2024-US-Presidential-Election)."
date: today
date-format: long
abstract: "In this paper, we develop a linear model to predict the outcome of the 2024 U.S. presidential election using a \"poll-of-polls\" approach that aggregates data from multiple well-known pollsters. By analyzing state-level polling trends, our model predicts that Kamala Harris is likely to secure key states, leading to an estimated electoral vote count of 341 for Harris and 197 for Donald Trump. This approach, which combines multiple polls, reduces bias and improves reliability, providing a more accurate picture of public sentiment leading up to the election."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

The structure of the paper is organized as follows: following @sec-intro, @sec-data presents the data collection and cleaning process, along with an overview of the variables used in the analysis. @sec-model introduces the forecasting models, explaining why the selected models are suitable for predicting election outcomes based on aggregated polling data. Then, @sec-result provides the results, highlighting key trends and predictions. Eventually, @sec-discussion concludes with a discussion of the findings, evaluating the reliability of the forecasts and identifying potential limitations of the models.

@@

ggplot(cleaned_data, aes(x = end_date, y = pct)) +
  geom_point(alpha = 0.5) +  # Add points for individual poll data
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +  # Show month and year, with one label per month
  labs(
    x = "Date of Poll",
    y = "Polling Percentage (%)",
  ) +
  theme_minimal() +  # Use a clean minimal theme
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),  # Center and adjust title size
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for readability
  ) +
  geom_smooth()
```

Using package `ggplot2` [@citeggplot], @fig-trump-pct shows Donald Trump's polling percentages over time from July to October 2024. Each black dot represents an individual poll conducted on a specific date, indicating the percentage of voters who supported Trump in that poll. The blue line represents a trendline, smoothing out the individual poll results to show the general trajectory of Trump's support over time.

@@

@@

The model assumes that the distribution of support percentage, given these polling characteristics, follows a normal distribution. This Gaussian assumption facilitates parameter estimation, which is a standard approach in linear regression. To prevent overfitting and ensure interpretability, we assume moderate priors, maintaining balanced uncertainty across our predictors. This approach enables us to assess the impact of polling characteristics on candidate support while ensuring stability in our findings. Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

The linear regression model built by our training dataset, using 242 data points, estimated the factors influencing Trump's support levels. For brevity, @tbl-summary-model only shows the first ten rows of the model’s coefficients, with the full model summary available in [Appendix -@sec-model-details]. The intercept is estimated at 44.224, representing the baseline support level when all predictors are at their reference levels. The model achieved an $R^2$ value of 0.803, indicating that 80.3% of the variance in polling outcomes is explained by the predictors included in the model. The adjusted $R^2$ value of 0.753 confirms that the model captures the relationships between variables without significant overfitting.

@@

Other variables, like `transparency_score` (-0.220) and `duration` (0.028), have smaller effects. The negative coefficient for `transparency_score` suggests a slight decrease in predicted support as transparency increases, potentially reflecting a sensitivity among the public to the transparency level in polling methods. Meanwhile, `sample_size` (0.002) shows a small positive effect, indicating that larger sample sizes modestly increase predicted support. A complete overview of all predictors is provided in [Appendix -@sec-model-details].

@@

# Create and display the results table
kable(missing_states, 
      col.names = c("State", "Electoral Votes", "Winner"), 
      digits = 2)
```

@@

state_predictions <- state_predictions %>%
  select(state, electoral_votes, winner)

final_predictions <- full_join(state_predictions, missing_states, by = c("state", "electoral_votes", "winner"))

@@

# Display the final table with added styling
kable(final_outcome, col.names = c("Candidate", "Final Votes"), digits = 2) %>%
  row_spec(1, background = "#FFCCCC") %>%    # Add light orange background for Trump row
  row_spec(2, background = "#C1D7FF") %>%    # Add light blue background for Harris row
  column_spec(2, bold = TRUE)
```

@@

# Plot the map using usmap and ggplot2
plot_usmap(data = final_predictions, values = "winner_color", regions = "states") +
  scale_fill_identity(
    name = "Predicted Winner",
    labels = c("Harris", "Trump"),
    guide = "legend"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8),
    plot.background = element_rect(fill = "white", color = NA)
  )
```

By using package `usmap` [@citeusmap], @fig-statemap displays the predicted outcome for each state in the 2024 U.S. Presidential Election, with red representing states where Trump is expected to win and blue for those favoring Harris.

@@

# Model details {#sec-model-details}

## Model summary

@tbl-summary presents the coefficients from our model analyzing factors that influence polling outcomes. The intercept value, 44.178, represents the baseline level of support when all other predictors are at their reference or baseline levels. Key variables include "numeric_grade," with a positive coefficient indicating that higher grades correlate with increased support. Different polling methodologies show varying effects; for example, "Mixed Voting" has a positive coefficient of 1.697, while "Online Panel" and "Probability Panel" methods have negative coefficients, suggesting that these methods might be associated with lower support levels in polling outcomes.

```{r}
#| label: tbl-summary
#| tbl-cap: "Coefficients from a regression model examining factors influencing polling outcomes, including polling methodology, duration, transparency, and state-level indicators."
#| echo: false
#| warning: false

modelsummary(
  list(
    "coefficient" = model
  )
)
```

## Posterior predictive check

In @fig-ppcheck, we implement a posterior predictive check, which displays the overlap between the observed data (denoted by $y$) and the replicated data generated from the model (denoted by $y_{rep}$). In a well-fitting model, the distribution of these replicated data should resemble the distribution of the observed data. Here, the replicated lines closely follow the shape and central tendency of the observed data’s density, especially around the peak. This similarity indicates that the model is capturing the overall pattern of the data reasonably well, suggesting a good fit. Deviations between the replicated and observed lines would indicate areas where the model might not be accurately capturing the data structure.

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheck
#| fig-cap: "Comparison of Observed Data Density (dark line) and Model-Generated Replications (light lines) shows close alignment, suggesting the model captures the main distributional characteristics of the data."

pp_check(model) +
  theme_classic() +
  theme(legend.position = "bottom")
```

## Diagnostics

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-rhat
#| fig-cap: "All parameters have R-hat values at or below 1.05, indicating strong convergence of the MCMC chains and reliable parameter estimates."

plot(model, "rhat")
```

@fig-rhat is a visualization of the Gelman-Rubin diagnostic, often denoted as R-hat, which assesses the convergence of a Bayesian model’s Markov Chain Monte Carlo (MCMC) samples. It shows the values of for various parameters in the model. An value close to 1 indicates good convergence, meaning the chains for each parameter have mixed well and are sampling from the same posterior distribution. In @fig-rhat, all parameters have an value at or below 1.05, suggesting that the model has achieved adequate convergence and the estimates are reliable.
